Clinical Language Encoding with ModernBERT

### 2025-04-15 article/article.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CLEM-ICD - Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler Cross},
  pdfkeywords={ModernBERT, natural language
processing, MIMIC-IV, ICD-10},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{CLEM-ICD - Clinical Language Encoding with
ModernBERT \\\large{Using ModernBERT to Improve Automated ICD-10
Classification} }
\author[1]{Tyler Cross%
\corref{cor1}%
\fnref{fn1}}
 \ead{tyler.cross@berkeley.com}

\affiliation[1]{organization={University of California, Berkeley, School
of Information},addressline={102 South
Hall},city={Berkeley},postcode={94720},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{Final project submission for the UC Berkeley Master of
Information and Data Science program, DATASCI 266: Natural Language
Processing with Deep Learning}

\begin{abstract}
Medical coding is a critical yet complex process that requires
translating clinical documentation into standardized ICD-10 codes.
Currently, healthcare providers rely on dedicated medical coders who
manually review extensive clinical notes to assign appropriate codes
from 150,000+ possible diagnostic and procedural codes. This project
builds upon the work of \citet{edin2024explainable}{]} with two key
innovations: (1) replacing RoBERTa with ModernBERT
\citep{warner2024modernbert} as the encoder to potentially improve
performance through its enhanced context window, and (2) adding a
LLM-powered reasoning layer that provides natural language explanations
for code predictions, bridging the gap between model predictions and
human validation. Using the MIMIC-IV dataset, we implement and evaluate
this approach using standard metrics (precision, recall, F1) and
explanation quality.
\end{abstract}

\begin{keyword}
    ModernBERT \sep natural language processing \sep MIMIC-IV \sep 
    ICD-10
\end{keyword}
\end{frontmatter}

Medical coding is a critical yet complex process in healthcare
administration. The translation of clinical documentation into
standardized ICD-10 codes is essential for healthcare reimbursement,
clinical research, and public health statistics. Currently, healthcare
providers rely on dedicated medical coders who manually review extensive
clinical notes to assign appropriate codes from over 150,000 possible
diagnostic and procedural codes, creating significant administrative
overhead.

For every \$1 of revenue collected by a hospital, approximately \$0.25
is spent on administrative tasks necessary to collect it. Non-doctor
workers outnumber doctors in healthcare roughly 16 to 1, with medical
coding representing a significant portion of this burden.

This project aims to improve automated medical coding by leveraging
ModernBERT's enhanced context window capabilities, potentially reducing
the administrative burden while maintaining or improving coding
accuracy.

\section{Background}\label{background}

\citet{edin2024explainable} recently demonstrated success in medical
coding using a RoBERTa-based approach. Their work showed promising
results but left room for improvement in both performance and contextual
understanding. Other notable contributions in this field include
{[}additional related work to be filled in{]}.

ModernBERT \citep{warner2024modernbert} represents an advancement over
previous BERT-like models, offering an enhanced context window of 8192
tokens compared to the 512 tokens in traditional models. This increased
context capacity is particularly valuable for medical documents, which
tend to be lengthy and contain important information distributed
throughout the text.

\subsection{Using CSL}\label{using-csl}

\section{Methods}\label{methods}

This project utilizes the MIMIC-IV dataset, a large, freely available
database comprising de-identified health data associated with hospital
stays. The dataset includes detailed clinical notes, diagnostic codes,
procedural information, and other health-related data from real hospital
encounters.

\subsection{System Architecture}\label{system-architecture}

Our approach replaces the RoBERTa model used in previous work
\citep{edin2024explainable} with ModernBERT
\citep{warner2024modernbert}, leveraging its enhanced context window to
better process lengthy clinical notes. We hypothesize that this will
lead to improved code prediction by enabling the model to capture more
comprehensive context from medical documents.

\subsection{Implementation Details}\label{implementation-details}

{[}To be completed with specific implementation details{]}

\subsection{Evaluation Approach}\label{evaluation-approach}

We evaluate our system using standard performance metrics: precision,
recall, F1-score, and accuracy compared to gold-standard human coding,
following the evaluation protocol established in previous work.

\section{Results and Discussion}\label{results-and-discussion}

{[}This section will be completed after obtaining experimental
results{]}

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler M Cross},
  pdfkeywords={clinical natural language
processing, ModernBERT, MIMIC-IV, MIMIC-III, MDACE, multi-label
classification},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{A Preprint }
\title{Clinical Language Encoding with ModernBERT}
\def\asep{\\\\\\ } % default: all authors on same column
\author{\textbf{Tyler M Cross}~\orcidlink{0009-0003-3529-8222}\\Master's
in Data Science\\University of California, Berkeley School of
Information\\Berkeley,
CA,\ 94720\\\href{mailto:tyler.cross@berkeley.edu}{tyler.cross@berkeley.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Medical coding---assigning standardized codes to clinical
documentation---is a costly, labor-intensive process. Automating this
task via multi-label classification faces challenges like long documents
and extreme class imbalance. We present CLEM-ICD, replacing the
RoBERTa/BERT base models common in prior frameworks with ModernBERT,
leveraging its 8192-token context window to better process lengthy
clinical notes. ModernBERT's optimized architecture (\textasciitilde138M
parameters) enhances computational efficiency and simplifies
implementation compared to prior approaches requiring complex
modifications for shorter-context models. Evaluating on the MDACE
dataset, CLEM-ICD achieves a Micro-F1 score of 47.6\%, outperforming
recent results by Edin et al. (2024) (41.9\% Micro-F1). On the standard
MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of
16.5\%, significantly outperforming the current state-of-the-art on
uncommon classes. CLEM-ICD demonstrates how modern architectural
advancements, particularly large context windows applied to encoder-only
models, can yield strong performance on complex tasks like automated
medical coding. We release our code and models to foster further
research.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
clinical natural language
processing \sep ModernBERT \sep MIMIC-IV \sep MIMIC-III \sep MDACE \sep 
multi-label classification

\section{Introduction}\label{sec-intro}

Medical coding, the assignment of standardized codes to clinical
documentation, is an essential step for medical billing and
reimbursement. The administrative processes surrounding billing and
insurance-related activities, which heavily rely on accurate coding,
represent a significant burden, consuming up to 25\% of healthcare
expenditures in the United States (Kocher and Sahni 2011; Tseng et al.
2018). Research has focused on attempts to automate coding by training
multi-label classification models, with transformer-based architectures
emerging as particularly effective (Huang, Tsai, and Chen 2022; Liu et
al. 2024). The PLM-ICD framework (Huang, Tsai, and Chen 2022), utilizing
BERT-based language models to encode clinical documents, established a
strong benchmark for automated ICD coding by effectively balancing model
complexity and predictive accuracy on the MIMIC-III dataset (A. E.
Johnson et al. 2016).

Despite advances, significant challenges persist in automated medical
coding systems. Clinical documents often exceed the standard context
windows (e.g., 512 tokens) of many transformer models, causing potential
information loss during truncation or requiring complex workarounds like
the segment pooling employed by PLM-ICD (Huang, Tsai, and Chen 2022).
Furthermore, the extreme class imbalance inherent in ICD coding hinders
performance on rare but clinically significant codes. Although
transformer-based approaches generally outperform older convolutional
and recurrent architectures, these limitations highlight the need for
architectures with natively expanded context windows, a capability
offered by modern transformer variants like ModernBERT (Warner et al.
2024), which can process lengthy clinical documents more holistically.

\section{Methods}\label{sec-methods}

We followed the established data preparation pipeline developed by Cheng
et al. (2023) for processing the MIMIC-III and MIMIC-IV datasets (A. E.
W. Johnson et al. 2023), heavily relying on the open-source code
provided by Edin et al. (2024) to match clinical visit notes with their
associated diagnostic and procedure codes. Specifically, for the results
compared against Edin et al. (2024) in Table~\ref{tbl-results}, we
utilized the MIMIC-III dataset (A. E. Johnson et al. 2016), focusing on
inpatient discharge summaries and their associated ICD-9 codes,
processed according to the splits defined by Cheng et al. (2023). These
notes originate from various clinicians across intensive care units at
the Beth Israel Deaconess Medical Center. While previous approaches like
PLM-ICD (Huang, Tsai, and Chen 2022) and its iterations employed
specialized architectures for clinical text encoding, we diverged from
this approach by adopting the more generic multi-label classification
architecture recommended by ModernBERT (Warner et al. 2024). This
decision simplified the architecture, facilitating experimentation and
fully leveraging ModernBERT's 8192-token context window.

Training transformer models with such extended context windows presented
significant computational challenges, particularly on consumer-grade
hardware. To address memory constraints, we implemented a combination of
gradient checkpointing during backpropagation and Flash Attention 2.0
(Dao et al. 2022). These optimizations enabled us to train our models on
a standard consumer GPU without degrading performance. To facilitate
direct comparison with existing approaches, we attempted to maintain the
same evaluation metrics employed by PLM-ICD and subsequent works (Huang,
Tsai, and Chen 2022; Edin et al. 2024; Liu et al. 2024), including micro
and macro-averaged F1 scores, precision, and recall metrics, which
represent the standard evaluation framework in this field.

\section{Results}\label{sec-results}

Our proposed CLEM-ICD model, leveraging the ModernBERT architecture with
an 8192-token context window, was evaluated on the MDACE dataset using
the standard multi-label classification metrics. The best performing
model achieved a Micro-averaged F1 score (Micro-F1) of 0.476,
Micro-Precision of 0.680, and Micro-Recall of 0.366 on the test set. The
Macro-averaged F1 score (Macro-F1) reached 0.038.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.4875}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1875}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1750}}@{}}
\caption{Comparison of multi-label classification performance metrics
between our CLEM-ICD model and the AttInGrad (TM) baseline from Edin et
al. (2024).}\label{tbl-results}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Model
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Precision (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Recall (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Micro-F1 (\%)
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Model
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Precision (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Recall (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Micro-F1 (\%)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
AttInGrad (TM) (Edin et al. 2024) & 40.2 \(\pm\) 3.0 & 43.9 \(\pm\) 4.8
& 41.9 \(\pm\) 3.4 \\
CLEM-ICD (Ours) & 68.0 & 36.6 & 47.6 \\
\end{longtable}

These results demonstrate a notable improvement over recent benchmarks.
For instance, comparing our Micro-F1 score of 47.6\% to the 41.9\%
reported by Edin et al. (2024) for their AttInGrad model on the same
MIMIC-III/MDACE benchmark dataset, our approach shows enhanced
performance. A key difference is that the Edin et al. (2024) results are
averaged over 10 runs, providing confidence intervals, whereas our
CLEM-ICD results are based on a single training run due to computational
constraints. Nevertheless, the CLEM-ICD model appears competitive with,
and potentially surpasses, state-of-the-art methods like PLM-ICD (Huang,
Tsai, and Chen 2022) under similar evaluation conditions, particularly
benefiting from the extended context capacity. The low Macro-F1 score
(3.8\% on this dataset), however, aligns with observations in prior
work, indicating persistent challenges in accurately classifying less
frequent codes within the highly imbalanced ICD code distribution.

To provide a direct comparison on the widely used MIMIC-III benchmark
dataset (A. E. Johnson et al. 2016), we evaluate CLEM-ICD against the
original PLM-ICD (Huang, Tsai, and Chen 2022) and subsequent
improvements reported by Liu et al. (2024). The results are summarized
in Table~\ref{tbl-mimic3-comparison}. Our CLEM-ICD model demonstrates a
strong Macro-F1 score, suggesting better performance on less frequent
codes compared to prior work, although the Micro-F1 score is lower in
this specific run.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.2523}}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}@{}}
\caption{Comparison of results on the MIMIC-III full test set (\%).
CLEM-ICD results are from a single run. P@k metrics were not computed
for CLEM-ICD in this run.}\label{tbl-mimic3-comparison}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Model
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Macro-F1 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Micro-F1 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
P@5 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
P@8 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
P@15 (\%)
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Model
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Macro-F1 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Micro-F1 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
P@5 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
P@8 (\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
P@15 (\%)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
PLM-ICD (Huang, Tsai, and Chen 2022) & 10.4 & 59.8 & 84.4 & 77.1 &
61.3 \\
BL-5 (Liu et al. 2024) & 11.1 \(\pm\) 0.1 & 60.7 \(\pm\) 0.1 & 85.2
\(\pm\) 0.2 & 78.0 \(\pm\) 0.2 & 62.4 \(\pm\) 0.1 \\
CLEM-ICD (Ours) & 16.5 & 54.6 & - & - & - \\
\end{longtable}

Note that while Macro-F1 and Micro-F1 scores are directly comparable,
the P@k metrics reported by Huang, Tsai, and Chen (2022) and Liu et al.
(2024) differ from the overall micro-averaged precision (65.6\%) and
recall (46.8\%) metrics recorded for our CLEM-ICD run.

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_recall_mdace.png}}

}

\subcaption{\label{fig-recall-mdace}Evaluation Recall (MDACE)}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_precision_mdace.png}}

}

\subcaption{\label{fig-precision-mdace}Evaluation Precision (MDACE)}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_macro_f1_mdace.png}}

}

\subcaption{\label{fig-macro-f1-mdace}Evaluation Macro-F1 Score (MDACE)}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_micro_f1_mdace.png}}

}

\subcaption{\label{fig-micro-f1-mdace}Evaluation Micro-F1 Score (MDACE)}

\end{minipage}%

\caption{\label{fig-learning-curves-mdace}Learning curves for CLEM-ICD
trained on MIMIC-III Inpatient Discharge Summaries (MDACE splits) for 10
epochs.}

\end{figure}%

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_recall_full.png}}

}

\subcaption{\label{fig-recall-full}Evaluation Recall (MIMIC-III Full)}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_precision_full.png}}

}

\subcaption{\label{fig-precision-full}Evaluation Precision (MIMIC-III
Full)}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_macro_f1_full.png}}

}

\subcaption{\label{fig-macro-f1-full}Evaluation Macro-F1 Score
(MIMIC-III Full)}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_micro_f1_full.png}}

}

\subcaption{\label{fig-micro-f1-full}Evaluation Micro-F1 Score
(MIMIC-III Full)}

\end{minipage}%

\caption{\label{fig-learning-curves-full}Learning curves for CLEM-ICD
trained on the full MIMIC-III dataset (PLM-ICD splits) for 20 epochs.}

\end{figure}%

Observing the learning curves (Figure~\ref{fig-learning-curves-mdace}
and Figure~\ref{fig-learning-curves-full}), the model trained on the
full MIMIC-III dataset (Figure~\ref{fig-learning-curves-full}) appears
to plateau relatively early in training, particularly for Micro-F1,
while the model trained on the smaller MDACE split
(Figure~\ref{fig-learning-curves-mdace}) shows more continued
improvement later into the 10 epochs. Both runs exhibited signs of
overfitting, suggesting that future training could benefit from
implementing early stopping based on validation set performance.

\section{Discussion}\label{discussion}

\subsection{Architectural
Considerations}\label{architectural-considerations}

Our results underscore the advantage of leveraging transformer
architectures with natively long context windows for processing lengthy
clinical narratives in automated ICD coding. ModernBERT's 8192-token
capacity allows CLEM-ICD to process entire documents holistically,
potentially avoiding context fragmentation issues that can arise from
the segment pooling techniques employed by prior work like PLM-ICD
(Huang, Tsai, and Chen 2022) to handle shorter context limits. This
architectural choice significantly simplifies the implementation,
relying on standard Hugging Face library components
(\texttt{AutoModelForSequenceClassification}) rather than requiring
bespoke modules for segmentation or label-specific attention mechanisms,
which were critical for PLM-ICD's performance. While this standard
classification head might be considered less sophisticated than
specialized approaches for extreme multi-label classification (Chang et
al. 2019; Liu et al. 2024), the strong representational power of
ModernBERT combined with its ability to access the full document context
appears to compensate effectively for this task.

\subsection{Base Model Choice and
Efficiency}\label{base-model-choice-and-efficiency}

The selection of ModernBERT-base (Warner et al. 2024) as the foundation
model provides benefits beyond its extended context window.
Incorporating architectural optimizations adapted from recent
decoder-only models, ModernBERT-base (\textasciitilde138M parameters) is
designed for improved computational efficiency, demonstrating enhanced
inference speed compared to previous BERT-style encoders, particularly
for variable-length inputs and long sequences due to its alternating
attention mechanism (Warner et al. 2024). This increased efficiency
holds potential for deployment in settings with limited computational
resources or for enabling faster processing of large clinical datasets.
However, the base model utilized here underwent general-domain
pretraining (including text, code, and scientific literature) without
specific adaptation to clinical or biomedical language. Consequently,
further performance improvements might be realized through future
domain-specific adaptation, either via continued pretraining on clinical
corpora or fine-tuning biomedically-specialized ModernBERT variants.

\subsection{Performance Analysis and
Limitations}\label{performance-analysis-and-limitations}

The effectiveness of this approach is reflected in the competitive
results. CLEM-ICD achieves a superior Micro-F1 on the MDACE benchmark
compared to AttInGrad (Edin et al. 2024) and demonstrates a markedly
improved Macro-F1 score on the full MIMIC-III dataset compared to both
PLM-ICD (Huang, Tsai, and Chen 2022) and the more recent BL-5 model (Liu
et al. 2024). This strong Macro-F1 performance suggests that the model's
access to the full, unsegmented context and potentially the broader
knowledge within ModernBERT's pretraining data aids significantly in
classifying less frequent codes, a persistent challenge in this domain.
This capability is particularly relevant for developing coder assistance
tools, as human coders often need more support with rare codes than
common ones.
